Experimental and numerical studies of the effects of resonant and random excitations of the proton beam in the LHC, with applications to the design of a pulsed hollow electron lens for active halo control

### 2018-02-27 resex.tex
Restored resex.tex and .gitignore
Intermediate step in including Miriam's bibliography; resex-bibdesk.bib and resex-mf.bib are still separate.
Revised figure and table in experimental setup (Section IV). Split setup section into experimental (Section IV) and simulations (Section V).

\documentclass[%
 reprint,
%superscriptaddress,
%groupedaddress,
%unsortedaddress,
%runinaddress,
%frontmatterverbose, 
%preprint,
%showpacs,preprintnumbers,
%nofootinbib,
%nobibnotes,
%bibnotes,
 amsmath,amssymb,
 aps,
%pra,
%prb,
%rmp,
prstab,
%prstper,
%floatfix,
]{revtex4-1}

@@

In this paper, we concentrate on the experimental and numerical
assessment of possible detrimental effects on the beam core of a
pulsed electron lens. Section~\ref{sec:hel} gives an introduction to
the concept of HELs and summarizes the design parameters of the HL-LHC
HELs. Section~\ref{sec:core} is dedicated to the sources of residual
fields from the HEL in the core region. Section~\ref{sec:exp}
describes the setup of simulations and experimental
conditions. Results and comparisons between simulations and
measurements are given in Section~\ref{sec:simex}, with discussion and
summary in Section~\ref{sec:sum}. Auxiliary data and plots are
collected in the Appendices.

@@

\section{Experimental and simulation setup\label{sec:exp}}
% put picture on top

\begin{table*}
	\caption{\label{tab:md_param}%
		Beam parameters and machine configuration as used in the two resonant excitation experiments in 2016 and~2017~\cite{resexmd2016,resexmd2017}. The plane of the excitation is abbreviated with H for horizontal, V for vertical and H+V horizontal and vertical at the same time.
	}
	\begin{ruledtabular}
		\begin{tabular}{lcc}
			Parameter & Experiment 2016 & Experiment 2017  \\
			\colrule
			beam &\multicolumn{2}{c}{Beam~1} \\
			beam energy &\multicolumn{2}{c}{injection energy, 450 GeV} \\\hline
			single bunch intensity &\multicolumn{2}{c}{$0.7\times10^{11}$} \\
			normalized emittance &\multicolumn{2}{c}{$2.5-3.5~\mathrm{\mu m}$} \\
			$4\sigma$ bunch length & \multicolumn{2}{c}{1.3~ns}\\
			$1\sigma$ bunch length & \multicolumn{2}{c}{9.7~cm}\\
			number of bunches & $12\times4=48$ single bunches & $3\times72=216$ bunches\\
			&  & (+ 1 pilot + 12 nominal) \\\hline
			injection optics, $\beta^*=11$~m & standard optics 2016 & standard optics 2017\\
			Landau damping octupoles  & \multicolumn{2}{c}{$I_{\mathrm{MO}}=+19.6~\mathrm{A}$ for MOF circuit and}\\
			& \multicolumn{2}{c}{$I_{\mathrm{MO}}=-19.6~\mathrm{A}$ for MOD circuit (standard 2016 settings)}\\\hline
			working point $(Q_x,Q_y)$ & (64.28,59.31) & (62.27,60.295) \\
			chromaticity $(Q'_x,Q'_y)$ & \multicolumn{2}{c}{(+15,+15)}\\\hline
			pulsing patterns  &$7^{\mathrm{th}}$~turn H; &$7^{\mathrm{th}}$~turn H,V,H+V; \\
			& $10^{\mathrm{th}}$~turn V; & $8^{\mathrm{th}}$~turn H,V,H+V; \\
			& &  random  H,V,H+V;\\
		\end{tabular}
	\end{ruledtabular}
\end{table*}

\begin{figure*}
	\begin{minipage}[t]{1.0\linewidth}
		\centering
		2016 experiment\\
		\includegraphics[width=0.9\linewidth]{bunchfilling_2016.png}	
	\end{minipage}
	\begin{minipage}[t]{1.0\linewidth}
		\centering
		$ $\\
		2017 experiment\\
		\includegraphics[width=0.9\linewidth]{bunchfilling_2017.png}	
	\end{minipage}
	\caption{\label{fig:fill} Filling and excitation scheme for LHC experiments in 2016 (top) and 2017 (bottom). In 2016 a total of 48~bunches was used and in 2017 a total of 216~bunches. Each bunch is indicated with a cyan vertical bar. The bunches were grouped in sub-groups of 4~for 2016 and 6~for 2017 bunches, where each sub-group experienced the same excitation pattern and amplitude. The excitation amplitude is shown in blue and red. In 2016 the excitation was only applied in one plane. In 2017 more bunches could be injected without compromising machine protection making it possible to test all three planes in one fill.}
\end{figure*}
%put picture on top

@@

The resonant excitation experiment in 2017 involved simultaneous measurements on three groups of 72 bunches with a dedicated excitation pattern and transverse feedback configuration on each sub-group of 6 bunches. In 2016 a similar configuration was used, but with only 48 bunches in total and sub-groups of 4 bunches. Both schemes are illustrated in Fig.~\ref{fig:fill}. In order to reduce the error of the measurements, the observables are averaged over each sub-group of bunches experiencing the same excitation amplitude, yielding in total 5~different amplitudes $n\Delta A$ with $n=1,\cdots, 5$ plus the references bunch without excitation for each transverse feedback configuration and excitation plane. Furthermore, the experiments were usually split in 3 different parts:
\begin{description}
	\item[Part 1] No excitation is applied in order to allow the distribution to reach an equilibrium state after injection.
	\item[Part 2] The excitation is applied with a first maximum excitation amplitude of $A_{\mathrm{max,1}}=5\Delta A_1$.
	\item[Part 3] The maximum excitation amplitude is further increased to $A_{\mathrm{max,1}} < A_{\mathrm{max,2}}=5\Delta A_2$.
\end{description}
Each part lasted approximately 10-15~minutes, which is considered to be a long enough time window to allow the beam distribution to reach its new equilibrium state. In the plots of the experimental results shown later in the paper, the maximum excitation amplitude is indicated with black horizontal arrows, while the average of each sub-group of bunches with the same excitation amplitude $n\Delta A$ is distinguished with different colors.

@@

\subsection{Overview of simulations\label{sec:sim}}
For the preparation and interpretation of the experiments two different types of simulations were performed:
\begin{enumerate}
	\item tracking of a Gaussian distribution, referred to as ``distribution tracking" in this paper, to obtain loss rates and emittance growth,
	\item Frequency map analysis (FMA) for the visualization of resonances~\cite{fmalaskar}.
\end{enumerate}
For both simulation types, the tracking code Lifetrac \cite{lifetrac} was used. The simulation parameters are summarized in Table~\ref{tab:sim_param}. Further details on the simulations are given in Refs.~\cite{md_sim_hel_res_ex_fitterer,resexmd2017}.
\begin{table}[h]
	\caption{\label{tab:sim_param}%
		Summary of simulation parameters for the distribution tracking and FMA analysis. For details see \cite{md_sim_hel_res_ex_fitterer,resexmd2017}.
	}
	\begin{ruledtabular}
		\begin{tabular}{lcc}
			Parameter & distribution tracking & FMA \\
			\colrule
			beam &\multicolumn{2}{c}{Beam~1} \\
			beam energy &\multicolumn{2}{c}{450 GeV} \\
			emittance & $3.5~\mu$m& $2.5~\mu$m\\
			$4\sigma$ bunch length & \multicolumn{2}{c}{1.3~ns}\\
			$1\sigma$ bunch length & \multicolumn{2}{c}{9.7~cm}\\
			particle & 6D Gaussian & equally spaced grid\\
			distribution &  distribution & in $x,y$ up to $10~\sigma$\\
			&  ($10^4$ particles) & with $\frac{\Delta p}{p_0}=0$\\
			turns tracked & $10^6$ turns & $10^4$ turns \\\hline
			optics & \multicolumn{2}{c}{2016 or 2017 injection optics,}\\
			& \multicolumn{2}{c}{$\beta_{\mathrm{IP1/5}}^*=11$~m}\\
			machine  & standard errors & no errors \\
			imperfections &with $a_1=b_1=0$\footnote{Orbit errors are disabled due to different implementation of the $a_1,b_1$ errors in Lifetrac and MAD-X. $b_2$ errors are adjusted to yield an average (over 60 seeds) peak $\beta$-beat of 15\% as expected from optics measurements in the LHC.} &  \\
			octupoles  & \multicolumn{2}{c}{$I_{\mathrm{MOF}}=+19.6~\mathrm{A}$, $I_{\mathrm{MOD}}=-19.6~\mathrm{A}$}\\
			tune $(Q_x,Q_y)$ & \multicolumn{2}{c}{(64.28,59.31) for 2016,}\\
			& \multicolumn{2}{c}{(62.27,60.295) for 2017}\\
			chromaticity & \multicolumn{2}{c}{(+15,+15)}\\
			$(Q'_x,Q'_y)$ & & \\\hline
			transv. aperture & $5.7~\sigma$ & - \\
			long. aperture & $10~\sigma$ & - \\
		\end{tabular}
	\end{ruledtabular}
\end{table}

\section{Experimental and simulation results\label{sec:simex}}
In the following we will jump forth and back between both experimental and simulation results as the experimental results are best understood on the base of the simulations. In summary, the most efficient pulsing patterns could be successfully predicted in simulations, which we will demonstrate on the example of the 2016 experiments (see Sec.~\ref{sec:pattern}). Afterwards we will present the simulation and experimental results of the tested excitation patterns, explicitly the $10^{\mathrm{th}}$~turn, $7^{\mathrm{th}}$~turn and $8^{\mathrm{th}}$~turn pulsing in Secs.~\ref{sec:simex10}--\ref{sec:simex8} as well as the random excitation in Sec.~\ref{sec:simexran}. As a last chapter, Sec.~\ref{sec:damp}, we will address the influence of the transverse damper in the presence of an external excitation source on emittance, losses and beam distribution.

>>>>

\documentclass[%
 reprint,
%superscriptaddress,
%groupedaddress,
%unsortedaddress,
%runinaddress,
%frontmatterverbose, 
%preprint,
%showpacs,preprintnumbers,
%nofootinbib,
%nobibnotes,
%bibnotes,
 amsmath,amssymb,
 aps,
%pra,
%prb,
%rmp,
prstab,
%prstper,
%floatfix,
longbibliography
]{revtex4-1}

@@

In this paper, we concentrate on the experimental and numerical
assessment of possible detrimental effects on the beam core of a
pulsed electron lens. Section~\ref{sec:hel} gives an introduction to
the concept of HELs and summarizes the design parameters of the HL-LHC
HELs. Section~\ref{sec:core} is dedicated to the sources of residual
fields from the HEL in the core region. Sections~\ref{sec:exp}
and~\ref{sec:sim} describe the experimental conditions and the setup
of the numerical tracking simulations. Results and comparisons between
simulations and measurements are given in Section~\ref{sec:simex},
with discussion and summary in Section~\ref{sec:sum}. Auxiliary data
and plots are collected in the Appendices.

@@

\section{Experimental setup}
\label{sec:exp}

\begin{table*}
  \caption{Beam parameters and machine configuration for the two
    resonant excitation experiments of~2016
    and~2017~\cite{resexmd2016, resexmd2017}. The plane of the
    excitation is abbreviated as H for horizontal, V for vertical and
    H+V for horizontal and vertical at the same time.}
  \label{tab:md_param}
  \begin{ruledtabular}
    \begin{tabular}{lcc}
      Parameter & Experiment 2016 & Experiment 2017  \\
      \colrule
      beam &\multicolumn{2}{c}{Beam~1} \\
      beam energy &\multicolumn{2}{c}{injection energy, 450 GeV} \\\hline
      single bunch intensity &\multicolumn{2}{c}{$0.7\times10^{11}$} \\
      normalized emittance &\multicolumn{2}{c}{$2.5$--\q{3.5}{\mu m}} \\
      $4\sigma$ bunch length & \multicolumn{2}{c}{1.3~ns}\\
      $1\sigma$ bunch length & \multicolumn{2}{c}{9.7~cm}\\
      number of bunches & $12 \times 4 = 48$ & $3 \times 72 = 216$ \\
                &  & (+ 1 pilot + 12 nominal) \\\hline
      injection optics, $\beta^* = \q{11}{m}$ & standard optics 2016 & standard optics 2017\\
      Landau-damping octupoles  & \multicolumn{2}{c}{$I_{\mathrm{MO}}
                                  = \q{+19.6}{A}$ for MOF circuit and}\\
                & \multicolumn{2}{c}{$I_{\mathrm{MO}} = \q{-19.6}{A}$
                  for MOD circuit (standard 2016 settings)} \\ \hline
      working point $(Q_x, Q_y)$ & (64.28, 59.31) & (62.27, 60.295) \\
      chromaticity $(Q'_x, Q'_y)$ & \multicolumn{2}{c}{(+15, +15)}\\ \hline
      pulsing patterns  & $7^{\mathrm{th}}$~turn H
                                  &$7^{\mathrm{th}}$~turn H, V, H+V \\
                & $10^{\mathrm{th}}$~turn V & $8^{\mathrm{th}}$~turn
                                              H, V, H+V \\
                & &  random  H, V, H+V\\
    \end{tabular}
  \end{ruledtabular}
\end{table*}

\begin{figure*}
  \begin{tabular}{c}
    \multicolumn{1}{l}{\emph{2016 experiment}} \\
    \includegraphics[width=0.95\textwidth]{bunchfilling_2016.png} \\
    \\
    \multicolumn{1}{l}{\emph{2017 experiment}} \\
    \includegraphics[width=0.95\textwidth]{bunchfilling_2017.png}
  \end{tabular}
  \caption{Bunch filling scheme and excitation patterns for the
    2016~(top) and 2017~(bottom) LHC experiments. In 2016, a total of
    48~bunches was used, whereas in 2017 there were 216~bunches. Each
    bunch is represented by a vertical cyan bar. The bunches were
    grouped in subsets of~4 in 2016 and in subsets of~6 in 2017. Each
    subset experienced the same excitation pattern and amplitude. The
    excitation amplitudes and relative phases are shown in blue or
    red. In 2016, the excitation was only applied in one plane. In
    2017, more injected bunches were allowed without compromising
    machine protection, so it was possible to test all~3 excitation
    planes in the same fill. The transverse damper was active on half
    of the bunches, indicated by the black lines.}
  \label{fig:fill}
\end{figure*}

@@

The resonant excitation experiment in 2017 involved simultaneous measurements on three groups of 72 bunches with a dedicated excitation pattern and transverse feedback configuration on each subset of 6 bunches. In 2016 a similar configuration was used, but with only 48 bunches in total and subsets of 4 bunches. Both schemes are illustrated in Fig.~\ref{fig:fill}. In order to reduce the error of the measurements, the observables are averaged over each subset of bunches experiencing the same excitation amplitude, yielding in total 5~different amplitudes $n\Delta A$ with $n=1,\cdots, 5$ plus the references bunch without excitation for each transverse feedback configuration and excitation plane. Furthermore, the experiments were usually split in 3 different parts:
\begin{description}
	\item[Part 1] No excitation is applied in order to allow the distribution to reach an equilibrium state after injection.
	\item[Part 2] The excitation is applied with a first maximum excitation amplitude of $A_{\mathrm{max,1}}=5\Delta A_1$.
	\item[Part 3] The maximum excitation amplitude is further increased to $A_{\mathrm{max,1}} < A_{\mathrm{max,2}}=5\Delta A_2$.
\end{description}
Each part lasted approximately 10-15~minutes, which is considered to be a long enough time window to allow the beam distribution to reach its new equilibrium state. In the plots of the experimental results shown later in the paper, the maximum excitation amplitude is indicated with black horizontal arrows, while the average of each subset of bunches with the same excitation amplitude $n\Delta A$ is distinguished with different colors.

@@

\section{Numerical tracking simulations}
\label{sec:sim}

For the preparation and interpretation of the experiments two different types of simulations were performed:
\begin{enumerate}
	\item tracking of a Gaussian distribution, referred to as ``distribution tracking" in this paper, to obtain loss rates and emittance growth,
	\item Frequency map analysis (FMA) for the visualization of resonances~\cite{fmalaskar}.
\end{enumerate}
For both simulation types, the tracking code Lifetrac \cite{lifetrac} was used. The simulation parameters are summarized in Table~\ref{tab:sim_param}. Further details on the simulations are given in Refs.~\cite{md_sim_hel_res_ex_fitterer,resexmd2017}.
\begin{table}[h]
	\caption{\label{tab:sim_param}%
		Summary of simulation parameters for the distribution tracking and FMA analysis. For details see \cite{md_sim_hel_res_ex_fitterer,resexmd2017}.
	}
	\begin{ruledtabular}
		\begin{tabular}{lcc}
			Parameter & distribution tracking & FMA \\
			\colrule
			beam &\multicolumn{2}{c}{Beam~1} \\
			beam energy &\multicolumn{2}{c}{450 GeV} \\
			emittance & $3.5~\mu$m& $2.5~\mu$m\\
			$4\sigma$ bunch length & \multicolumn{2}{c}{1.3~ns}\\
			$1\sigma$ bunch length & \multicolumn{2}{c}{9.7~cm}\\
			particle & 6D Gaussian & equally spaced grid\\
			distribution &  distribution & in $x,y$ up to $10~\sigma$\\
			&  ($10^4$ particles) & with $\frac{\Delta p}{p_0}=0$\\
			turns tracked & $10^6$ turns & $10^4$ turns \\\hline
			optics & \multicolumn{2}{c}{2016 or 2017 injection optics,}\\
			& \multicolumn{2}{c}{$\beta_{\mathrm{IP1/5}}^*=11$~m}\\
			machine  & standard errors & no errors \\
			imperfections &with $a_1=b_1=0$\footnote{Orbit errors are disabled due to different implementation of the $a_1,b_1$ errors in Lifetrac and MAD-X. $b_2$ errors are adjusted to yield an average (over 60 seeds) peak $\beta$-beat of 15\% as expected from optics measurements in the LHC.} &  \\
			octupoles  & \multicolumn{2}{c}{$I_{\mathrm{MOF}}=+19.6~\mathrm{A}$, $I_{\mathrm{MOD}}=-19.6~\mathrm{A}$}\\
			tune $(Q_x,Q_y)$ & \multicolumn{2}{c}{(64.28,59.31) for 2016,}\\
			& \multicolumn{2}{c}{(62.27,60.295) for 2017}\\
			chromaticity & \multicolumn{2}{c}{(+15,+15)}\\
			$(Q'_x,Q'_y)$ & & \\\hline
			transv. aperture & $5.7~\sigma$ & - \\
			long. aperture & $10~\sigma$ & - \\
		\end{tabular}
	\end{ruledtabular}
\end{table}

\section{Results}
\label{sec:simex}

In the following we will jump forth and back between both experimental and simulation results as the experimental results are best understood on the base of the simulations. In summary, the most efficient pulsing patterns could be successfully predicted in simulations, which we will demonstrate on the example of the 2016 experiments (see Sec.~\ref{sec:pattern}). Afterwards we will present the simulation and experimental results of the tested excitation patterns, explicitly the $10^{\mathrm{th}}$~turn, $7^{\mathrm{th}}$~turn and $8^{\mathrm{th}}$~turn pulsing in Secs.~\ref{sec:simex10}--\ref{sec:simex8} as well as the random excitation in Sec.~\ref{sec:simexran}. As a last chapter, Sec.~\ref{sec:damp}, we will address the influence of the transverse damper in the presence of an external excitation source on emittance, losses and beam distribution.
